The yaq Project: \\ Standardized Software Enabling Flexible Instrumentation

### 2022-11-17 paper.tex
DDK edits

\yaq{} introduces a concept called ``Traits'' which are collections of related messages that are shared among multiple protocols.
Motors implement the ``\texttt{has-position}'' trait, which defines ``\texttt{set\_position}'', ``\texttt{get\_position}'', and ``\texttt{get\_units}''.
Sensors would implement the ``\texttt{is-sensor}'' trait which defines ``\texttt{get\_measured}'', ``\texttt{get\_channel\_names}'', and ``\texttt{get\_channel\_units}''.
Protocols which implement a trait must support all of the messages from the trait.
Importantly, specific protocols can also implement arbitrary additional messages that are not defined by any trait.

@@

The second hardware interface barrier: Certain specialty hardware have inconvenient interface requirements.
Since \yaq{} enables multiple machines, any hardware requirements can be addressed by putting a machine for that specific hardware on the network.
The Raspberry Pi which drives several stepper motors can be placed onto a private network to communicate with the primary instrument computer using \yaq{}.
Because each \yaq{} daemon is running in its own process, the software environment can be tailored to its needs.
A client running up to date Python 3.10 communicates seamlessly with a daemon running Python 3.7.

The third hardware interface barrier: Some hardware interfaces are blocking.
Monolithic orchestration software often necessitates separate threads for each component hardware interface, a fragile pattern in which small mistakes become both critical and elusive errors.
In \yaq{}, each daemon is only responsible for a single hardware interface.
%Each \yaq{} daemon runs as a separate process (possibly not even on the same machine) and thus many of the hard concurrency problems with communicating with hardware are mitigated.
It is expected that each message call over the \yaq{} interface will return rapidly, ensuring that client applications are not blocked for extended periods of time.
This principle applies even when the message starts an action that might take several seconds to complete, such as homing a motor or initiating a measurement for a sensor.
In these instances, the initial message simply starts the action and returns, with a separate message provided to retrieve results when they exist.
In order to know how long to wait, the ``is-daemon`` trait provides a message called ``is\_busy'' which should return "true" while the long running action is not complete, and "false" once it is finished.
Additionally, multiple clients can communicate with the same daemon simultaneously.
A complex instrument may involve multiple operators watching sensor data in real time, while one program is orchestrating the hardware and recording the data.

@@

Existing experimental orchestration software is often highly inflexible.
An experimentalist will spend many hours in lab repeating acquisitions because it is too challenging to add repetition functionality to their software.
A laser lab needs to spend weeks on software development when introducing a single new step into their experimental procedure.
Researchers are disappointed to realize that they are forced to start from scratch when developing software for a similar instrument built with trivially different hardware.

@@

Unique experiments will always need custom orchestration and user experience.
We believe that novel instrumentation development naturally and necessarily includes the creation of targeted software.
Developing experimental software is an iterative process tied to the scientific goals of the instrument.
Often experimentalists must apply their specialty scientific knowledge to develop this software.\cite{SegalJudith2005a} % citation does not compile
\yaq{} is architected to encourage better software development practices when creating such programs.
In a \yaq{} context, orchestration code and graphical control interfaces are implemented as clients.
These clients are automatically simpler because they only need to implement the \yaq{} standard and do not need to include the vast array of hardware-specific communication interfaces.
Beyond this, clients can use traits to interact with similar hardware identically.
A client written to perform a two dimensional fluorescence experiment using two Acton monochromators will also work with Horiba monochromators without any modification.

>>>>

\yaq{} introduces a concept called ``traits'' which are collections of related messages that are shared among multiple protocols.
Motors implement the ``\texttt{has-position}'' trait, which defines ``\texttt{set\_position}'', ``\texttt{get\_position}'', and ``\texttt{get\_units}''.
Sensors would implement the ``\texttt{is-sensor}'' trait which defines ``\texttt{get\_measured}'', ``\texttt{get\_channel\_names}'', and ``\texttt{get\_channel\_units}''.
Protocols which implement a trait must support all of the messages from the trait.
Importantly, specific protocols can also implement arbitrary additional messages that are not defined by any trait.

@@

The second hardware interface barrier: Certain specialty hardware have inconvenient interface requirements.
Since \yaq{} enables multiple machines, any hardware requirements can be addressed by putting a machine for that specific hardware on the network.
The Raspberry Pi which drives several stepper motors can be placed onto a private network to communicate with the primary instrument computer using \yaq{}.
Because each \yaq{} daemon is running in its own process, the software environment can be tailored to its needs.
A client running up-to-date Python 3.11 communicates seamlessly with a daemon running Python 3.7.

The third hardware interface barrier: Some hardware interfaces are blocking.
Monolithic orchestration software often necessitates separate threads for each component hardware interface, a fragile pattern in which small mistakes become both critical and elusive errors.
In \yaq{}, each daemon is only responsible for a single hardware interface.
%Each \yaq{} daemon runs as a separate process (possibly not even on the same machine) and thus many of the hard concurrency problems with communicating with hardware are mitigated.
It is expected that each message call over the \yaq{} interface will return rapidly, ensuring that client applications are not blocked for extended periods of time.
This principle applies even when the message starts an action that might take several seconds to complete, such as homing a motor or initiating a measurement for a sensor.
In these instances, the initial message simply starts the action and returns, with a separate message provided to retrieve results when they exist.
In order to know how long to wait, the ``is-daemon`` trait provides a message called ``is\_busy'', which returns "true" while the long running action is not complete, and "false" once it is finished.
Additionally, multiple clients can communicate with the same daemon simultaneously.
A complex instrument may involve multiple operators watching sensor data in real time, while one program is orchestrating the hardware and recording the data.

@@

Existing experimental orchestration software is often highly inflexible.
An experimentalist will spend many hours in lab manually repeating acquisitions because it is too challenging to add repetition functionality to their software.
A laser lab needs to spend weeks on software development when introducing a single new step into their experimental procedure.
Researchers are disappointed to realize that they are forced to start from scratch when developing software for a similar instrument built with trivially different hardware.

Unique experiments will always need custom orchestration and user experience.
We believe that novel instrumentation development naturally and necessarily includes the creation of targeted software.
Developing experimental software is an iterative process tied to the scientific goals of the instrument.
Often experimentalists must apply their specialty scientific knowledge to develop this software.\cite{SegalJudith2005a} % citation does not compile

@@

\yaq{} is architected to encourage better software development practices when creating such programs.
In a \yaq{} context, orchestration code and graphical control interfaces are implemented as clients.
These clients are automatically simpler because they only need to implement the \yaq{} standard and do not need to include the vast array of hardware-specific communication interfaces.
Beyond this, clients can use traits to interact with similar hardware identically.
A client written to perform a two dimensional fluorescence experiment using two Acton monochromators will also work with Horiba monochromators without any modification.
